Appendix A: Stochastic simulations of noisy phenomena

### 2018-02-07 appendixA/appendixA.Rmd
WIP

```{r include=FALSE}
## note: needs library Cairo
knitr::opts_chunk$set(message=FALSE, cache = 1, warning = FALSE,  dev="CairoPDF")
```

```{r libraries, cache=FALSE}
library(nimble)
library(RcppRoll)
library(regimeshifts) ## devtools::install_github("cboettig/regimeshifts")
library(adaptivetau)
library(tidyverse)
library(ggthemes)
library(gridExtra)
ggplot2::theme_set(theme_bw())
colours <- ptol_pal()(2)
```

@@

extrafont::loadfonts()
ggplot2::theme_set(hrbrthemes::theme_ipsum_rc())
```

@@

Levins' Patch model as an individual-based birth-death process:

@@

We simulate this using the Gillespie's exact stochastic simulation algorithm (SSA), as implemented in the `adaptivetau` R package.  We declare the transition events: birth increases the state $n$ to $n+1$, and death $n$ to $n-1$, and describe the rates associated with them:

@@

sim_small <- adaptivetau::ssa.exact(init.values = c(n = 50),
                       transitions, rateF,
                       params = list(c=c, e=e, N=100),
                       tf=30) %>% as_tibble()

sim_large <- adaptivetau::ssa.exact(init.values = c(n = 500),
                       transitions, rateF,
                       params = list(c=c, e=e, N=1000),
                       tf=30) %>% as_tibble()
```

We will also create a separate data frame (`tibble`) corresponding to our theoretical predictions from the system size expansion:

@@

```{r}
gillespie <- bind_rows("large"=sim_large, 
                       "small"=sim_small,  
                       .id = "system_size") %>% 
  bind_rows(theory)
write_csv(gillespie, "gillespie.csv")
```

@@

Simulations of the quasi-cycles can easily be performed in NIMBLE, which allows us to express the stochastic models in the familiar BUGS syntax.  Change the values of the model and noise parameters below to explore these results.

```{r}

quasicycle  <- nimble::nimbleCode({

  x[1] <- x0
  y[1] <- y0

  for(t in 1:(N-1)){
    mu_x[t] <- x[t] + x[t] * r * (1 - x[t] / K) - b * x[t] * y[t]
    x[t+1] ~ dnorm(mu_x[t], sd = sigma_x)
    mu_y[t] <- y[t] + c * x[t] * y[t] - d * y[t]
    y[t+1] ~ dnorm(mu_y[t], sd = sigma_y)
  }

p <- 
  data.frame(
  data.frame(r = .1, K = 5,  b = .1, c = .1, d = .1, N = 800),
  data.frame(sigma_x = c(.00001,0.01),  sigma_y = c(.00001,0.01)))

f <- function(constants){
  model <- compileNimble(nimbleModel(quasicycle, 
                                     constants = constants, 
                                     inits = list(x0 = 1, y0 = 1)))
  set.seed(123)
  simulate(model)
  tibble(t = seq_along(model$x), 
         x = model$x, 
         y = model$y, 
         sigma = constants$sigma_x)
}

quasicycle_df <- p %>% rowwise() %>% do(f(.))        
write_csv(quasicycle_df, "quasicycles.csv")
```

@@

```{r quasicycles, fig.width=7}
read_csv("quasicycles.csv") %>%
  gather(species, population, -t, -sigma) %>%
  mutate(sigma = as.factor(sigma)) %>%
  mutate(sigma = recode(sigma, "1e-05" = "A.", "0.01" = "B.")) %>%
  rename(time = t) %>%
  ggplot(aes(time, population, col = species)) + geom_line() + 
  facet_wrap(~ sigma, ncol=2) + scale_color_ptol() 
```

@@

```{r}
theory <- 
  tibble(x= seq(0,2, length.out = 100)) %>%
  mutate(g = x * p$r * (1 - x / p$K), 
         c = p$a * x ^ p$Q / (x^p$Q + p$H^p$Q)) %>%
  mutate(potential = - cumsum(g - c)) %>%
  gather(curve, y, -x, -potential)

@@

```{r}
max_days <- 1000

z <- y <- x <- numeric(max_days)
z[1] <- y[1] <- x[1] <- 7e4 # 1.7e5
set.seed(123)
for(day in 1:(max_days-1)){
  x[day+1] <- dai(x[day], epsilon = rnorm(1,0, 0.001), DF = 1790)
  y[day+1] <- dai(y[day], epsilon = rnorm(1,0, 0.001), DF=1799)
  z[day+1] <- dai(z[day], epsilon = rnorm(1,0, 0.001), DF = 1800)

@@

df <- data.frame(t = 1:max_days, x = x, y = y, z = z) %>%
  gather(series, pop, -t)

ggplot(df, aes(t, pop, col=series)) + 
  geom_line()

@@

We are now ready to create the panels shown in Figre 4:

@@

```{r eval=FALSE}
p <- list(r = .4, K = 1, sigma = .15, N = 500)
equib <- (1 + sqrt(1 - 8 * p$sigma^2/p$K^2) )/ 2

logistic  <- nimble::nimbleCode({

  x[1] <- x0
  for(t in 1:(N-1)){
    mu[t] <- x[t] + x[t] * r * (1 - x[t] / K)
    x[t+1] ~ dnorm(mu[t], sd = sigma)
  }

model <- 
nimbleModel(logistic,
            constants = p,
            inits = list(x0 = 1))
cmodel <- compileNimble(model)
set.seed(123)

@@

```{r}
read_csv("inflation.csv") %>% 
  gather(statistic, value, -t) %>%
  ggplot(aes(t,value, col=statistic, lty=statistic)) + 
 # ggplot(aes(t,ave)) +
  geom_line(lwd=1) + 
  scale_color_ptol() + 
  coord_cartesian(xlim=c(100,500), ylim=c(1.2, .5))
```

>>>>

```{r include=FALSE}
## note: needs library Cairo
knitr::opts_chunk$set(message=FALSE, cache = 1, warning = FALSE,  dev="cairo_pdf")
```

```{r libraries, cache=FALSE}
library(nimble)
library(RcppRoll)
## devtools::install_github("cboettig/regimeshifts")
library(regimeshifts) 
library(adaptivetau)  ## For Gillespie simulation only
library(tidyverse)
library(ggthemes)
library(gridExtra)

## Set a default color pallete
colours <- ptol_pal()(2)
```

@@

## One-time setup
hrbrthemes::import_roboto_condensed()   # One-time use to set up fonts 
#d <- read.csv(extrafont:::fonttable_file(), stringsAsFactors = FALSE)
#d[grepl("Light", d$FontName), ]$FamilyName <- "Roboto Condensed Light"
#write.csv(d, extrafont:::fonttable_file(), row.names = FALSE)

## load fonts (needed each time)
extrafont::loadfonts()
ggplot2::theme_set(hrbrthemes::theme_ipsum_rc())
```

@@

Levins' Patch model [@Levins1969] as an individual-based birth-death process:

@@

We simulate this using the Gillespie's exact stochastic simulation algorithm (SSA) [@Gillespie1977; see overview in @Hastings2012], as implemented in the `adaptivetau` R package.  We declare the transition events: birth increases the state $n$ to $n+1$, and death $n$ to $n-1$, and describe the rates associated with them:

@@

## Run simulations
sim_small <- adaptivetau::ssa.exact(init.values = c(n = 50),
                       transitions, rateF,
                       params = list(c=c, e=e, N=100),
                       tf=30) %>% as_tibble()

sim_large <- adaptivetau::ssa.exact(init.values = c(n = 500),
                       transitions, rateF,
                       params = list(c = c, e = e, N = 1000),
                       tf=30) %>% as_tibble()
```

We will also create a separate data frame (`tibble`) corresponding to our theoretical predictions from the system size expansion.  We use popular functions from the `tidyverse` suite of R packages for the manipulation of `data.frames`.  Below, the `mutate` function is used to add new columns for plus and minuse one standard deviation based on theoretical calculation of $\sqrt{N\tfrac{e}{c}}$, and the `map_dfr` function is used to apply this to both `small` and `large` cases.  We use `tidyverse` syntax  [see @Ross2017] instead of base R in these examples for the following reasons:

- the resulting code is more concise, due to use of higher-level relational data abstractions
- the resulting code is closer to natural semantics.  This is in part because consecutive steps are connected by pipes `%>%`rather than temporary variable assignment, and in part by the use of functions named by verbs corresponding to standard relational data concepts.
- `tidyverse` is widely used and well documented, see <http://r4ds.had.co.nz/> for an accessible and thorough introduction.

@@

```{r}
## combine data.frames, using column called "system_size"
## to indicate large noise or small noise simulations
gillespie <- 
  bind_rows("large" = sim_large, 
            "small" = sim_small,  
            .id = "system_size") %>% 
  bind_rows(theory)

## Write data to file
write_csv(gillespie, "gillespie.csv")
```

@@

Simulations of the quasi-cycles can easily be performed in NIMBLE, which allows us to express the stochastic models in the familiar BUGS syntax.  Change the values of the model and noise parameters below to explore these results.  While the models below can easily be implemented in base R, the BUGS syntax for expressing a stochastic model provides a more generic formulation that can easily be used in parameter estimation as well; see @nimble.  This can also imporve computational performance because @nimble compiles these models into C++ code.

```{r}
## Define the BUGS model:
quasicycle  <- nimble::nimbleCode({
  x[1] <- x0
  y[1] <- y0
  for(t in 1:(N-1)){
    mu_x[t] <- x[t] + x[t] * r * (1 - x[t] / K) - b * x[t] * y[t]
    x[t+1] ~ dnorm(mu_x[t], sd = sigma_x)
    mu_y[t] <- y[t] + c * x[t] * y[t] - d * y[t]
    y[t+1] ~ dnorm(mu_y[t], sd = sigma_y)
  }
})

## creates a data.frame of parameter combinations for each noise level
p <- 
  data.frame(
  data.frame(r = .1, K = 5,  b = .1, c = .1, d = .1, N = 800),
  data.frame(sigma_x = c(.00001,0.01),  sigma_y = c(.00001,0.01)))

## define a function to perform a simulation given a set of parameters (constants)
f <- function(constants){
  model <- compileNimble(nimbleModel(quasicycle, 
                                     constants = constants, 
                                     inits = list(x0 = 1, y0 = 1)))
  ## set random seed for reproducibility
  set.seed(123)
  ## Run the actual simulation, using nimble
  simulate(model)

  ## Assemble the results into a data.frame (tibble).  
  tibble(t = seq_along(model$x), 
         x = model$x, 
         y = model$y, 
         sigma = constants$sigma_x)
}

## Do this simulation for each set of parameters, given 
quasicycle_df <- p %>% rowwise() %>% do(f(.))

## Write results to file for later reference and plotting
write_csv(quasicycle_df, "quasicycles.csv")
```

@@

```{r quasicycles, fig.width=7}
read_csv("quasicycles.csv") %>%
  ## Data in long from
  gather(species, population, -t, -sigma) %>%
  ## Re-arrange and and re-lable for the plots:
  mutate(sigma = as.factor(sigma)) %>%
  mutate(sigma = recode(sigma, "1e-05" = "A.", "0.01" = "B.")) %>%
  rename(time = t) %>%
  ggplot(aes(time, population, col = species)) + geom_line() + 
  facet_wrap(~ sigma, ncol=2) + scale_color_ptol() 
```

@@

```{r}
theory <- 
  tibble(x= seq(0,2, length.out = 100)) %>%
  ## As before, we add columns with mutate
  mutate(g = x * p$r * (1 - x / p$K), 
         c = p$a * x ^ p$Q / (x^p$Q + p$H^p$Q)) %>%
  mutate(potential = - cumsum(g - c)) %>%
  ## assemble resulting data into "long" form with the
  ## column headings as named here:
  gather(curve, y, -x, -potential)

@@

```{r}
max_days <- 1000
## Perform three simultaneous simulations, x,y,z, using
## different dilution factors DF

## Intialize variables
z <- y <- x <- numeric(max_days)
z[1] <- y[1] <- x[1] <- 7e4 # 1.7e5

## set random seed for reproduciblity
set.seed(123)

## simulate for max_days
for(day in 1:(max_days-1)){
  x[day+1] <- dai(x[day], epsilon = rnorm(1,0, 0.001), DF = 1790)
  y[day+1] <- dai(y[day], epsilon = rnorm(1,0, 0.001), DF=1799)
  z[day+1] <- dai(z[day], epsilon = rnorm(1,0, 0.001), DF = 1800)

@@

## gather results into long data.frame 
df <- data.frame(t = 1:max_days, x = x, y = y, z = z) %>%
  gather(series, pop, -t)

## and plot results
ggplot(df, aes(t, pop, col=series)) + 
  geom_line()

@@

We are now ready to create the panels shown in Figure 4:

@@

```{r eval=FALSE}

## Parameters
p <- list(r = .4, K = 1, sigma = .15, N = 500)
equib <- (1 + sqrt(1 - 8 * p$sigma^2/p$K^2) )/ 2

## BUGS model
logistic  <- nimble::nimbleCode({
  x[1] <- x0
  for(t in 1:(N-1)){
    mu[t] <- x[t] + x[t] * r * (1 - x[t] / K)
    x[t+1] ~ dnorm(mu[t], sd = sigma)
  }
})

## Compile and run model
model <- 
nimbleModel(logistic,
            constants = p,
            inits = list(x0 = 1))
cmodel <- compileNimble(model)
set.seed(123)

@@

```{r}
read_csv("inflation.csv") %>% 
  gather(statistic, value, -t) %>%
  ggplot(aes(t,value, col=statistic, lty=statistic)) + 
  geom_line(lwd=1) + 
  scale_color_ptol() + 
  coord_cartesian(xlim=c(100,500), ylim=c(1.2, .5))
```
